---
title: "Lab 4: LTER Network Data"
subtitle: 'ESS 330 - Quantitative Reasoning'
author:
  - name: "Zachary Cramton"
    url: "https://zcramton.github.io"
    email: "ZCramton@colostate.edu"
toc: true
execute:
  echo: true
format:
  html:
    self-contained: true
---

## Introduction to the Data
```{r}
# Install LTER Data Sampler
#remotes::install_github("lter/lterdatasampler")

# Install necessary libraries
library(tidyverse)
library(ggpubr)
library(lterdatasampler)
library(car)
library(visdat)
library(broom)
library(flextable)

# Retrieve the and_vertebrates data set
?and_vertebrates

# Explore data structure
str(and_vertebrates)
and_vertebrates %>% 
  glimpse() %>% 
  vis_dat()
```
### Part 1: Univariate and Bivariate Staistics
## Guided Data Analysis Part 1
```{r}
## Guided Data Analysis Part 1
and_vertebrates %>% 
  filter(species == "Cutthroat trout") %>% 
  drop_na(unittype) %>% 
  count(unittype)

trout_clean <- and_vertebrates %>% 
  filter(species == "Cutthroat trout") %>%    # Filter for cutthroat
  filter(unittype %in% c("C", "P", "SC")) %>%   # Filter for the 3 most abundant unittypes
  drop_na(unittype, section)   # Drop NA values for unittype and section

# Save Cutthroat trout table
cont_table <- table(trout_clean$section, trout_clean$unittype)

# Conduct chi-squared test on the Cutthroat trout data
chisq.test(cont_table)

# Plot the Cutthroat trout data

# Bar plot
trout_clean_barplot <- trout_clean %>% 
  count(unittype, section) %>% 
  ggbarplot(x = 'unittype', y = 'n',
            fill = 'section',
            palette = c("#00AFBB", "#E7B800"))

# Box Plot
trout_clean_boxplot <- trout_clean %>% 
  drop_na(weight_g) %>% 
  ggviolin(x = "section", y = "weight_g",
           add = "boxplot",
           color = "section",
           palette = c("#00AFBB", "#E7B800"))
  
# T-Test Assumptions
cc_weight <- trout_clean %>% 
  filter(section == "CC") %>% 
  pull(weight_g)

og_weight <- trout_clean %>%  
  filter(section == "OG") %>% 
  pull(weight_g)

var.test(cc_weight, og_weight)

# Plot Histograms
ggarrange(gghistogram(cc_weight, main = "Clear Cut"), gghistogram(og_weight, main = "Old Growth"))

# Test log normalization
var.test(log(cc_weight), log(og_weight))

# Default t-test with log-normalized data
t.test(log(trout_clean$weight_g) ~ trout_clean$section, var.equal = TRUE)

# Welch Two Sample t-test
t.test(trout_clean$weight_g ~ trout_clean$section, var.equal = FALSE)

# Coastal Giant Salamander
# Filter for salamander data
sally_clean <- and_vertebrates %>% 
  filter(species == "Coastal giant salamander") %>% 
  drop_na(length_2_mm, weight_g)   # Remove NA values from the data.

# Display salamander histograms
ggarrange(gghistogram(sally_clean$length_2_mm, title = "Length"),
gghistogram(sally_clean$weight_g, title = "Weight"))

# Test for normal distribution
s <- sally_clean  %>%  
  slice_sample(n = 5000) 

shapiro.test(s$length_2_mm)
shapiro.test(s$weight_g)

# Display log-normalized Salamander histograms
ggarrange(
 gghistogram(log(sally_clean$length_2_mm), title = "Length"), 
 gghistogram(log(sally_clean$weight_g), title = "Weight") )

# Correlation Test for Salamander Data
cor.test(log(sally_clean$length_2_mm), log(sally_clean$weight_g))

# Visualized correlation for Salamander Data
sally_clean %>%
  mutate(log_length = log(length_2_mm), log_weight = log(weight_g)) %>% 
  ggscatter(x = 'log_length',
            y = 'log_weight',
            alpha = .35,
            add = "loess")

# Spearman Correlation Test for Salamander Data
cor.test(sally_clean$length_2_mm, sally_clean$weight_g, method = "spearman")
```

# Part 1 Lab Exercise
```{r}
## Question 1-1: Conduct a chi-square test similar to the one carried out above, but test for a relationship between forest type (`section`) and channel unit (`unittype`) for Coastal giant salamander abundance.

# Filter and clean salamander data
sal_clean <- and_vertebrates %>% 
  filter(species == "Coastal giant salamander") %>%    # Filter for salamanders
  drop_na(unittype, section)   # Drop NA values for unittype and section

# Save Salamander contingency table
sal_cont_table <- table(sal_clean$section, sal_clean$unittype)


# Chi-Sq Test for Section and Unittype
sal_chisq_result <- chisq.test(sal_cont_table)

# Box Plot for Chi-sqaure test
sal_box_plot <- sal_clean %>% 
  count(unittype, section) %>% 
  ggpubr::ggbarplot(x = 'unittype', y = 'n',
                    fill = 'section',
                    palette = c("darkblue", "darkorange"),
                    add = "mean_se")
#ggsave("sal_chisq_test.png", plot = sal_box_plot, width = 10, height = 6)

# Print Chi-sq results
print(sal_chisq_result)
print(sal_box_plot)


## Question 1-2: Test the hypothesis that there is a significant difference in species biomass between clear cut and old growth forest types for the Coastal Giant salamander.


# Clean theadditional data
sal_clean <- sal_clean %>% 
  drop_na(weight_g)

# Equal Variance Test
sal_cc_weight <- sal_clean %>% 
  filter(section == "CC") %>% 
  pull(weight_g)

sal_og_weight <- sal_clean %>%  
  filter(section == "OG") %>% 
  pull(weight_g)

var.test(sal_cc_weight, sal_og_weight)

# Test for normal distribution
sal <- sal_clean %>%  
  slice_sample(n = 5000) 

# Visualizing data normality
ggarrange(gghistogram(sal_cc_weight, main = "Clear Cut"),
          ggpubr::gghistogram(sal_og_weight, main = "Old Growth"))

# Test log normalization
var.test(log(sal_cc_weight), log(sal_og_weight))

# Welch Two Sample t-test
t.test(sal_clean$weight_g ~ sal_clean$section, var.equal = FALSE)


## Question 1-3: Test the correlation between body length (snout to fork length) and body mass for Cutthroat trout. 

# Find the length variable (?and_vertebrates conducted in Chunk #1).
  # Variable is length_1_mm

# Clean the data
trout_clean <- and_vertebrates %>% 
  filter(species == "Cutthroat trout") %>% 
  drop_na(length_1_mm, weight_g)

# Look at distribution of variables
ggarrange(gghistogram(trout_clean$length_1_mm, title = "Length"),
          gghistogram(trout_clean$weight_g, title = "Weight"))

# Shapiro-Wilk normality test only runs for 5000 observations 
trout_sample <- trout_clean %>% 
  slice_sample(n = 5000)   # Adjust data frame to only contain 5000 obs.

# Run S-W Norm Test for both vars
shapiro.test(trout_sample$length_1_mm)
shapiro.test(trout_sample$weight_g)

# Log transform vizualization
ggarrange(
  gghistogram(log(trout_clean$length_1_mm), title = "Length"),
  gghistogram(log(trout_clean$weight_g), title = "Weight")
)

# Spearman Correlation Test
cor.test(trout_clean$length_1_mm, trout_clean$weight_g, method = "spearman")

```
### Question 1-1:
  The chi-squared = 200.71, df = 5, p-value is less than 0.05 (P < 2.2e-16), which means that there is a statistically significant relationship between forest type and channel unit in regards to coastal giant salamander abundance. This supports rejecting the null hypothesis.

### Question 1-2:
  The results of the variance test suggest that the variences are not equal. The p-value is less than 0.05 (p < 1.439e-07). This supports rejecting the null hypothesis that the variances are equal.
  Visualizing the normality of the data, it appears right skewed, suggesting that a log transformation should be used to normalize the data or that a Welsch t-test should be conducted. 
  Testing log normalization returned a p-value less than 0.05 (p < 5.299e-03) meaning even after adjustment we must reject the null hypothesis that the data is normally distrubuted. Given that the log normalization didn' work a Welsh t-test is required.
  The p-value for the Welch two sample t-test is less than 0.05 (p < 8.629e-07). This p-value supports rejecting the null hypothesis as the coastal giant salamander did not have the same weight in the two forest sections.
  
### Question 1-3:
  Both variables appear skewed (not normally distributed). Shapiro-Wilk normality test used to confirm. The p-value is less than 0.05 for both (p < 2.2e-16), so we reject the null hypothesis, meaning that our data does not fit a normal distribution. 
  The data can either be log-normalized or use the Pearson's correlation test or we can use the Spearman correlation test. The data appears to follow a non-normal distribution, requiring the Spearman correlation test.
  The p-value is less than 0.05 (p < 2.2e-16), we reject the null. This supports a significant, positive relationship between length and weight for the Cutthroat trout. There is also a very high correlation coefficient.
  
### Part 2: Multivariate Statistics
## Guided Data Analysis Part 2
```{r}
# Data set
data("pie_crab")

# Explore the data
glimpse(pie_crab)
vis_dat(pie_crab)
?pie_crab

# sample size per site
count(pie_crab, site)
summary(pie_crab)

# ANOVA
pie_crab |> 
  ggboxplot(x = 'site', y = 'size', col = 'site') +
  geom_jitter(size =.25) + 
  theme(legend.postition = "none")

# Assumptions 
# Normality
norms <- pie_crab |> 
  nest(data = -site) |>
  mutate(Shapiro = map(data, ~ shapiro.test(.x$size)),
         n = map_dbl(data, nrow),
         glance_shapiro = map(Shapiro, broom::glance)) |>
  unnest(glance_shapiro)

flextable::flextable(dplyr::select(norms, site, n, statistic, p.value)) |>
  flextable::set_caption("Shapiro-Wilk normality test for size at each site")

# Residuals
(res_aov <- aov(size ~ site, data = pie_crab))
gghistogram(res_aov$residuals)
shapiro.test(res_aov$residuals)

# Equal Variances
leveneTest(size ~ site, data = pie_crab)

# ANOVA Tests
# Welch's ANOVA
oneway.test(size ~ site, data = pie_crab, var.equal = FALSE)

# Filter a subset of the sites
pie_sites <- pie_crab |> 
  filter(site %in% c("GTM", "DB", "PIE"))

# Check for equal variance
leveneTest(size ~ site, data = pie_sites)

# Note that the variances are equal (p = 0.5802), so we can proceed with the ANOVA

# ANOVA for the data subset
pie_anova <- aov(size ~ site, data = pie_sites)

# View the ANOVA results 
summary(pie_anova)

# Post-hoc Tukey's HSD Test
TukeyHSD(pie_anova)

# Linear Regression
pie_lm <- lm(size ~ latitude, data = pie_crab)

# View the results of the linear model
summary(pie_lm)

# Linear regression visualized
pie_crab |> 
  ggscatter(x = 'latitude', y = 'size', 
            alpha = .35, 
            add = "reg.line")

# Predictions from linear regression
new_lat <- data.frame(latitude = c(32, 36, 38))
broom::augment(pie_lm, newdata = new_lat)

# Multiple linear regression
pie_mlm <- lm(size ~ latitude + air_temp + water_temp, data = pie_crab)

summary(pie_mlm)

# Test mlm for correlations
pie_crab |> 
  select(latitude, air_temp, water_temp) |> 
  cor()
```
## Part 2 Lab Exercise
```{r}
## Prework for Question 2-1
pie_crab %>% 
  ggboxplot(x = 'site', y = 'size', col = 'site') +
  geom_jitter(size =.25) +
  theme(legend.position = "none")

# Shapiro-Wilk Normality Test
norms <- pie_crab %>%  
  nest(data = -site) %>% 
  mutate(Shapiro = map(data, ~ shapiro.test(.x$size)),
         n = map_dbl(data, nrow),
         glance_shapiro = map(Shapiro, broom::glance)) %>%
  unnest(glance_shapiro)

flextable(select(norms, site, n, statistic, p.value)) %>% 
  set_caption("Shapiro-Wilk Normality Test for Size at Each Site")

# Normal distribution check
(res_aov <- aov(size ~ site, data = pie_crab))
gghistogram(res_aov$residuals)
shapiro.test(res_aov$residuals)

# Levene's Test
leveneTest(size ~ site, data = pie_crab)

# Welch's ANOVA
oneway.test(size ~ site, data = pie_crab, var.equal = FALSE)

# Filter a subset of the sites
pie_sites <- pie_crab |> 
  filter(site %in% c("GTM", "DB", "PIE"))

# Check for equal variance
leveneTest(size ~ site, data = pie_sites)

# Note that the variances are equal (p = 0.5802), so we can proceed with the ANOVA

# ANOVA for the data subset
pie_anova <- aov(size ~ site, data = pie_sites)

# View the ANOVA results 
summary(pie_anova)

# Post-hoc Tukey's HSD Test
TukeyHSD(pie_anova)

## Question 2-1
# Organize by latitude
pie_crab <- pie_crab %>% 
  arrange(latitude) %>% 
  mutate(site = factor(site, levels = unique(site)))

# Plot data
box_plot_crab_sites <- ggplot(pie_crab, aes(x = site, y = size)) +
  geom_boxplot(fill = "orange", color = "black") +
  geom_jitter(size =.25) +
  labs(title = "Carapace Width by Site, Ordered by Latitude",
       x = "Site",
       y = "Carapace Width (mm)") +
  theme_minimal()   
print(box_plot_crab_sites)

#ggsave("box_plot_crab_sites.png", plot = box_plot_crab_sites, width = 10, height = 6, dpi = 300)

## Question 2-2
# Linear regression model
pie_lm_water <- lm(size ~ water_temp_sd, data = pie_crab)
summary(pie_lm_water)

# Linear Regression Line of Best fit
LOBF <- pie_crab %>% 
  ggscatter(x = 'water_temp_sd', y = 'size',
  alpha = .35,
  add = "reg.line")
print(LOBF)
#ggsave("LOBF.png", plot = LOBF, width = 10, height = 6, dpi = 300)

## Question 2-3
# Check for correlations
pie_crab %>% 
  dplyr::select(latitude, air_temp_sd, water_temp_sd) %>% 
  cor()

# Multiple linear regression model
crab_mlm <- lm(size ~ latitude + air_temp_sd + water_temp_sd, data = pie_crab)
summary(crab_mlm)
```
### Question 2-1:
  The results of the ANOVA test are highly significant. At least one site has significantly different carapace widths than the other sites. We know this because the p-value is less than 0.05 (p < 2.2e-16), leading us to reject the null hypothesis.
  The Tukey's HSD Test shows us that the site relationships: GTM-DB, PIE-DB, and PIE-GTM are statistically significant, because their p-values are all less than 0.05. This represents that each of these related sites are significantly different from each other.
  
### Question 2-2:
  The p-value is greater than 0.05, we fail to reject the null hypothesis. There is not enough statistical evidence to represent a significant relationship between water_temp_sd and crab carapace width.

### Question 2-3:
  air_temp_sd and latitude have a higher correlation coefficient, with it being greater than 0.7 (0.79321301), but water_temp_sd and air_temp_sd, and latitude and water_temp_sd, are not highly correlated. Highly correlated variables increase the complexity of the linear regression model. 
  The overall p-value is less than 0.05 (p < 2.2e-16), suggesting a significant impact from the combination of predictors on crab carapace width. The individual p-values vary. Latitude is the only one less than 0.05, meaning it is the only statistically significant relationship with an effect on crab size. air_temp_sd and water_temp_sd both have p-values greater than 0.05, failing to reject the null hypothesis. There is not enough statistical evidence to represent a statistically significant relationship between these two predictors and crab carapace width.